Fused regression for multi-source gene regulatory network inference

### 2016-04-07 paper1.tex
hi
figure legeeends

\begin{equation}
\text{argmin}_{\beta^{S}} \displaystyle\sum_{S \in \{1, 2\}} ||X^{S}\beta^{S}(t) - Y^{S}||^2 + \lambda_R||\beta^{S}(t)||^2 + \lambda_S\displaystyle \sum_{\substack{(g,h) \in orth,\\
 (h,l) \in orth}} ||\beta^{S}_{g,k}(t) - \beta_{h,l}^{S}(t-1)||^2
\end{equation}
Note that this is almost identical to equation 5, but now the network $\beta$ is a function of the iteration number $t$. On each step, we compute $\beta$s that minimize a penalized cost function where the fusion penalties encourage similarity between a parameter and its fused-to parameter from the previous iteration's solution. This process is iterated until the estimated $\beta$s converge. Because each iteration reduces the error between $\beta(t)$ and $\beta(t-1)$, and because $\beta(t) = \beta(t-1)$ is the globally optimal solution, this process must eventually converge to the same network as equation \ref{eqn:main}. Although we have not produced bounds on the convergence rate, which also depends on the structure of constraints, in practice a small number of iterations are necessary.

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig1.pdf}
  \caption{\label{fusion} a. schematic representation of gene relationships b. We generate two networks with 10 TFs by 200 genes, with 75\% sparsity, 0.1 measurement noise, and 0.3 fusion noise. The x-axis represents the number of conditions drawn from species 1, and the y-axis tracks the number of conditions from species 2 used. Performance is mean square error of network 1; as the number of conditions from the first species increases, our ability to recover network 1 also improves. But if we increase the number of conditions from the related species 2, we are better able to learn the network describing species 1. c. We generate a series of networks with 20 TFs by 200 genes, with 50\% sparsity, 0.1 measurement noise, and 0.1 fusion noise. We increase amount of genes fused from 0 to 0.75, with 0.25 increments and evaluate performance on the whole network, interactions which have fusion constraints (constrained part of the network), and interactions which do not have fusion constraints (non-constrained), and the effect of increasing the fusion weight.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig2.pdf}
  \caption{\label{syntheticgrid} a. We generate a series of networks with 10 TFs by 200 genes, with 75\% sparsity, 0.1 measurement noise, and 0.1 fusion noise. We hold lamR constant at 1.0 and 2.0 respectively. The x-axis represents lamS and the y-axis represents the proportion of nodes which are orthologous. As the networks increase in similarity, performance improves and benefits from a large fusion penalty weight. We measure performance by showing ratio of mean squared error relative to MSE without fusion. b. Here is another set of networks of 10 TFs by 200 genes, with 75\% sparsity 0.1 measurement noise, and 0.75 fusion noise. We hold and lamR constant at 1.0 and 2.0 respectively and measure performance in the same way as previous figure.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig3.pdf}
  \caption{\label{adapt-schematic} a. Adaptive fusion is a quadratic around the origin, and tapers off and plateaus, allowing differences in fused interaction weights to grow if the expression data supports this, without incurring further penalty from the fusion. b. The adaptive fusion penalty is modified from SCAD (smoothly clipped absolute deviation) and MCP (minimax concave penalty) functions. It shares with these penalties the zero derivative further from the origin. }
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig4_2.png}
  \caption{\label{adaptivefusion} a.  Here we use two network of 35 TFs by 200 genes, with 0.4 of the genes and TFs fused, 0.1 measurement noise, and 0.1 fusion noise. We hold lamP and lamR constant at 1, and solve without fusion, using L2 fusion, and adaptive fusion, with lamS = 10 for both L2 and adaptive fusion. We plot the interaction weights; each point's coordinates represents the weight for network 1 and the weight for network 2. The solver was given a list of orthologs; the green points represent interactions between TFs and genes whose orthology information is correct, and the red points represent interactions between TFs and genes, where the orthology of one or both is incorrectly reported to the solver. We see that the L2 forces the points onto a diagonal, and that fused interactions adopt similar weights, regardless of the accuracy of the fusion. Our adaptive fusion solver is able to correctly identify and 'unfuse' incorrectly fused interactions. b. Here we plot the mean squared error of correctly fused and incorrectly fused (green and red dots, respectively) interactions for each solver.}
\end{center}
\end{figure}

>>>>

\begin{equation}
\argmin_{\beta^S} \displaystyle\sum_{S \in \{1, 2\}} ||X^{S}\beta^{S}(t) - Y^{S}||^2 + \lambda_R||\beta^{S}(t)||^2 + \lambda_S\displaystyle \sum_{\substack{(g,h) \in orth,\\
 (h,l) \in orth}} ||\beta^{S}_{g,k}(t) - \beta_{h,l}^{S}(t-1)||^2
\end{equation}
Note that this is almost identical to equation 5, but now the network $\beta$ is a function of the iteration number $t$. On each step, we compute $\beta$s that minimize a penalized cost function where the fusion penalties encourage similarity between a parameter and its fused-to parameter from the previous iteration's solution. This process is iterated until the estimated $\beta$s converge. Because each iteration reduces the error between $\beta(t)$ and $\beta(t-1)$, and because $\beta(t) = \beta(t-1)$ is the globally optimal solution, this process must eventually converge to the same network as equation \ref{eqn:main}. Although we have not produced bounds on the convergence rate, which also depends on the structure of constraints, in practice a small number of iterations are necessary.

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig1.pdf}
  \caption{\label{fusion} \textbf{A.} Schematic representation of the the generation of fusion constraints from orthology mappings. Dashed arrows indicate potential regulatory interactions, while solid arrows denote orthology. We introduce fusion constraints for pairs of interactions for which both the regulator and regulated gene are orthologs of one another. In this example, we would introduce a constraint between the $(A, B)$ and $(A', B')$ interactions and the $(A, C)$ and $(A', B')$ interactions. \textbf{B.} In order to demonstrate the utility of fused network inference in combining data, we generate two networks with 10 TFs and 200 genes (75\% sparsity). Mean squared error of the inferred vs. true coefficient matrices for network 1 are plotted as a function of the number of conditions generated for species 1 (x-axis) and the number of conditions generated for species 2 (y-axis). As expected, increasing the number of samples available for the species of interest improves network inference performance. However, because we are fusing to data from a related species, similar gains are observed when increasing the amount of data available in this second species. \textbf{C-F} Show the varying effects of fusion on simulated networks with different levels of conservation. We generate a series of networks with 20 TFs by 200 genes in two species, (50\% sparsity) while varying the fraction of gene orthologies in the simulated networks. For each network, we evaluated AUPR on one of the species for: all interactions (blue line), interactions with fusion constraints (green line), and interactions without fusion constraints (orange line). At every level of conservation, constrained interactions show the largest benefits of fusion, with the magnitude of the benefit growing with fraction of orthologous genes. When the networks are highly conserved, however, even interactions that are not directly constrained through fusion are recovered more accurately as $\lambda_S$ increases.
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig2.pdf}
  \caption{\label{syntheticgrid} This figure demonstrates the interaction between fusion weight ($\lambda_S$), the degree of orthology coverage of networks, and the degree of similarity between fused interactions. \textbf{A.} We generated a series of pairs of 10 TF by 200 gene networks (75\% sparsity) in two species. These networks had minimal fusion noise, so that pairs of interactions linking orthologs had nearly identical weight. We varied the fraction of genes with orthologs (y-axis) and the weight of fusion in network inference (x-axis), and measured performance as the mean-squared error of the true vs inferred network weights. In order to more clearly visualize the varying effect of fusion, performance is plotted in units of the of the unfused MSE (left column) for each level of orthology. This was necessary because each row represents a different pair of networks, generated with a different level of orthology, for which baseline performance varies. Performance gains from fusion are, as expected, largest when the degree of orthology is large. However, even when the fraction of genes with orthologs is relatively small (top row), we observe gains from fusion. Importantly, when fused interactions are nearly identical, larger fusion weights always outperform smaller fusion weights. \textbf{B.} Simulates the case where genes which are orthologs may have different regulatory weights. We generated a series of networks as in \textbf{A.}, but with larger (7.5\times) gaussian noise added the weights of fused interactions. As in \textbf{A.}, benefits from fusion were observed at every level of gene orthology. However, unlike \textbf{A.}, there was an optimal intermediate value of $\lambda_S$ that traded off between the benefits of fusion and the cost of combining heterogenous data.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig3.pdf}
  \caption{\label{adapt-schematic} Adaptive fusion loss function (\textbf{A}) and derivative of loss function (\textbf{B}). \textbf{A.} Adaptive fusion is a quadratic around the origin, begins to taper at $a/2$, and plateus at $a$. After the plateu, increasing the difference in interaction weight of fused interactions does not further affect the penalty incurred through fusion. As a result, interaction weights in this zone are effectively unfused from one another (the fusion penalty behaves like a constant). \textbf{B.} Shows the derivative of the adaptive fusion penalty, which is used to implement adaptive fusion through local quadratic approximation. The adaptive fusion penalty is modified from SCAD (smoothly clipped absolute deviation) and MCP (minimax concave penalty) functions. It shares with these penalties the zero derivative far from the origin. }
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{fig4_2.png}
  \caption{\label{adaptivefusion} In order to evaluate the performance of adaptive fusion in unfusing interactions that have diverged through evolution, we performed a series of simulations inferring networks given a partially false list of orthology mappings. These false orthologs were orthologs that were generated from random pairs of non-orthologous genes, and therefore did not reflect true similarity of regulatory interactions. Networks were generated with 35 TFs by 200 genes, 60\% orthology coverage and 40\% false orthology coverage. \textbf{A.} We plot the interaction weights between pairs of fused interactions in network 1 (x-axis) and network 2 (y-axis) following network inference without fusion ($\lambda_S=0$). Interactions that are generated from false orthologs, which do not reflect true similarity of regulatory weight, are marked as red, while interactions generated from true orthologs are shown in green. Because fused interactions in the simulated networks were, in reality, almost identical in regulatory weight, interactions with true fusion constraints should fall on the diagonal. However, because a very small number of samples (20 for each species) was provided to the network inference procedure, the inferred weights are only weakly correlated (green dots). As expected, false fusion constraints give rise to uncorrelated weights (red dots). \textbf{B.} When fit with fused-L2, fusion constraints give rise to very similar weights in the two species. This is desireable for the true fusion constraints, but introduces error for the false constraints. \textbf{C.} Adaptive fusion run on the same network unfuses constraints for which the inferred weights are dissimilar beyond a certain point. In this example, unfused interactions are almost entirely interactions between false orthologs. \textbf{D-F} Show the distribution of the absolute value of the difference in inferred weight for interactions with true fusion constraints (green) and false fusion constraints (orange)}
\end{center}
\end{figure}
